---
title: "TreeSHARP Introduction"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# TreeSHARP Overview

This vignette demonstrates how to simulation data and fit a DTM regression model with Stan using the `TreeSHARP` package.


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r source, message=FALSE, include=FALSE}
library(TreeSHARP)
```

## Defining a Tree

The user must specify a tree structure (via `ape`) for their data when fitting the model.
In this example, we use a simple tree with only 3 leaf nodes.


```{r tree}

my_tree <- read.tree(text = "(A, (B, C));")
plot(my_tree, "c", direction = "downward", main = "Example Tree")
nodelabels(bg="white")

```

## Simulating Data

We create data using `simulate_DTM`, which will produce an X matrix and coefficient values that will be used to generate the DTM response matrix, Y.

```{r sim-DTM}


sim_data <- simulate_DTM(subject_sim = 100,
                         num_cov = 2,
             subject_N = rep(50, 100),
             tree = my_tree,
             covariates_sim = 10,
             which_coninuous = 1:10,
             rho = 0.3, 
             seed = 220 )

```

## Fitting a Model 

To fit a model with `treesharp_stan`, three inputs are required: a covariate matrix, X; the response matrix, Y; and a tree.
Other `rstan` inputs, such as the number of chains, initial values, etc., may also be provided.
The `treesharp_stan` function will fit DTM regression model using a shared shrinkage horseshoe prior as recommended in our paper.
We also provide functions for the other two priors discussed in the paper: the original "unique shrinkage" horseshoe prior (`dtm_horseshoe_stan`) and independent normal (`dtm_normal_stan`).
Both functions also require X, Y, tree inputs, with `dtm_normal_stan` also requiring the user to provide the standard deviation to be used for the normal distribution.

```{r fit, eval = FALSE}

options(mc.cores = parallel::detectCores())

mod <- treesharp_stan(X = sim_data$X,
               Y = sim_data$Y,
               tree = my_tree,
               init = 0,
               chains = 4,
               seed = 1989,
               iter = 1000)

```
# Additional Functions

This package includes additional functions that may be helpful when fitting a DTM regression model or working with DTM data.

## Tree Counts

The function `DTM_counts` will take a $Y$ matrix that contains only leaf counts and return the corresponding counts for all leaf and internal nodes.
You may opt to have them ordered by branch index rather than node index or to have the counts returned as proportions.
Both of these options default to `FALSE`.

```{r ex-DTM_counts}

DTM_counts(Y = sim_data$Y[1:5,], tree = my_tree)

```

## DTM_likelihood

Calculates the (log-)likelihood for the data given the provided Dirichlet parameters (gamma) and tree structure.
The $y$ input should be provided in the format output by `DTM_counts` when `branches = TRUE`.
Optionally, you may specify whether to return it as a log-likelihood (default; recommended) and whether to provide a single likelihood value (`simplify = TRUE`; default) or provide them for each observation (`simplify = FALSE`).

```{r ex-DTM-likelihood}

g1 <- matrix(c(1, 1, 1, 1,
            1, 2, 2, 1,
            1, 2, 4, 5), byrow = TRUE,nrow=3)

y1 <- matrix(c(10, 2, 4,
               5, 4, 2,
               1, 0, 22), nrow = 3, byrow = TRUE)

yb <- DTM_counts(y1, my_tree)

DTM_likelihood(y = yb, gamma = g1, tree = my_tree)
DTM_likelihood(y = yb, gamma = g1, tree = my_tree,
               simplify = FALSE)

```


## Predict DTM

Produce the expected proportions for each node given the provided Dirichlet tree and parameters.
If $N_i$ is provided, it will return expected counts.

```{r ex-predict_DTM}

predict_DTM(gamma = g1, tree = my_tree, N_i = NULL)

predict_DTM(gamma = g1, tree = my_tree, N_i = c(100,20, 500))

```

## Sample DTM

Draw a random sample from a DTM distribution with the given tree, parameters, and size, N. The length of N should be equal to the number of rows in gamma.

```{r ex-sample_DTM}

sample_DTM(gamma = g1, N = c(100,100,43), tree= my_tree)

```

